---
title: "Multi-Row Headers"
author: "Lisa DeBruine"
date: "2018-02-22"
tags: "r, data import, headers"
---

[Download the Rmd notebook for this example](multirow_headers.Rmd)

```{r load-libraries, message = FALSE}
library(tidyverse)
```

A student on our help forum recently asked for help making a wide-format dataset long. When I tried to load the data, I realised the first three rows were all header rows. Here's the code I wrote to deal with it.

First, I'll make a small CSV "file" below. In a typical case, you'd read the data in from a file.
```{r data}
demo_csv <- "SUB1, SUB1, SUB1, SUB1, SUB2, SUB2, SUB2, SUB2, SUB3, SUB3, SUB3, SUB3
COND1, COND1, COND2, COND2, COND1, COND1, COND2, COND2, COND1, COND1, COND2, COND2
X, Y, X, Y, X, Y, X, Y, X, Y, X, Y
10, 15, 6, 2, 42, 4, 32, 5, 34, 54, 12, 19
4, 43, 7, 34, 56, 43, 2, 33, 29, 37, 2, 28
77, 12, 14, 75, 36, 85, 3, 2, 33, 44, 10, 9"
```

If you try to read in this data, you'll get an error message about the duplicate column names.
```{r get-data-bad, eval = FALSE}
data <- read_csv(demo_csv)
```
`Duplicated column names deduplicated: 'SUB1' => 'SUB1_1' [2], 'SUB1' => 'SUB1_2' [3]...`

Instead, you should read in just the header rows by setting `n_max` equal to the number of header rows and `col_names` to `FALSE`.

```{r get-headers}
data_head <- read_csv(demo_csv, n_max = 3, col_names = FALSE)
```

You will get a table that looks like this:

```{r, echo = F, results='asis'}
knitr::kable(data_head, "html") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"))
```

You can then transpose the table (rotate it) and make new header names by pasting together the names of the three headers.

```{r new-names}
new_names <- data_head %>%
  t() %>% # transposes the data and turns it into a matrix
  as_tibble() %>% # turn it back to a tibble
  mutate(name = paste(V1, V2, V3, sep = "_")) %>%
  pull(name)
```

Now you can read in the data without the three header rows. Use `skip` to skip the headers and set `col_names` to `FALSE`.

```{r get-data}
data <- read_csv(demo_csv, skip = 3, col_names = FALSE)
```

Then set the column names to the new names.

```{r set-new-names}
names(data) <- new_names
```

If your data are set up with multiple headers, you'll probably want to change them from wide to long format. Here's a quick example how to use `gather`, `separate`, and `spread` to do this with variable names like above.

```{r wide-to-long}
data_long <- data %>%
  # add row IDs if each row doesn't already have uniquely identifying column(s)
  mutate(row = 1:nrow(.)) %>% 
  # gather creates a column of variable names and a column of values
  gather("var", "val", new_names) %>% 
  # split the variable names into their three component parts
  separate(var, c("subID", "condition", "coord"), sep = "_") %>%
   # put X and Y in separate columns 
  spread(coord, val)
```

```{r, echo = F, results='asis'}
knitr::kable(data_long, "html") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"))
```

