---
title: "Plot Comparison"
author: "Lisa DeBruine"
date: "2018-03-28"
tags: "ggplot, ticyverse, dataviz"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r, messages = FALSE, warning=FALSE}
library(tidyverse)
```

I compared bar plots to violin plots in a recent talk to make the point that real data plotted with the full distribution make your effects look less impressive than minimalist bar charts that just show the means and standard errors, but give you a much better idea of what's going on with your data.

I thought I'd post a quick tutorial for anyone who wants to see some code for creating violin-box plots and split-violin plots.

First, let's simulate some data from a 2x2 design with a cross-over interaction with a 0.5 SD effect size.

```{r}

n <- 100
data <- tibble(
  sex = rep(c("male", "female"), n),
  face_sex = rep(c("male", "female"), each = n)
) %>%
  mutate(
    dv = rnorm(n*2, 0, 1),
    effect = ifelse(sex==face_sex, .5, 0),
    dv = dv + effect
  )
  
```

I like to create a theme for all the plots in a talk or paper. This one is my standard white-on-black talk theme.  

```{r}
bgcolor <- "black"
textcolor <- "white"
my_theme <- theme(
    plot.background = element_rect(fill = bgcolor, colour = bgcolor),
    panel.background = element_rect(fill = NA),
    legend.background = element_rect(fill = NA),
    legend.position="none",
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    text = element_text(family='Fira Code', colour = textcolor, size=20),
    axis.text = element_text(family='Fira Code', colour = textcolor, size=15)
  )
```

## Bar Plot

I've commented it out below, but I recommend alsways using `ggsave` to save your plots for papers or talks. They are much better resolution than the plots you copy out of an Rmarkdown notebook.

```{r plotcomp-bar}
data %>%
  group_by(sex, face_sex) %>%
  summarise(
    mean = mean(dv),
    se = sd(dv)/sqrt(n())
  ) %>%
  ggplot(aes(sex, mean, fill=face_sex)) +
  geom_hline(yintercept=0, color=textcolor, size=1) +
  geom_col(color = "white", position="dodge", alpha = 0.5) +
  geom_errorbar(aes(ymin = mean-se, ymax=mean+se), 
                width=0.1, 
                position=position_dodge(0.9), 
                color=textcolor) +
  ylab("DV") +
  xlab("Participant Sex") +
  scale_fill_manual(values = c("#3D66CC", "#892110")) +
  my_theme
  
#ggsave("bar.png", width=10, height = 6)
```

Notice how the bar plot hides the real range of the data. This is what it would look like plotted with the y-axis ranges shown below.
```{r plotcomp-bar-range}
data %>%
  group_by(sex, face_sex) %>%
  summarise(
    mean = mean(dv),
    se = sd(dv)/sqrt(n())
  ) %>%
  ggplot(aes(sex, mean, fill=face_sex)) +
  geom_hline(yintercept=0, color=textcolor, size=1) +
  geom_col(color = "white", position="dodge", alpha = 0.5) +
  geom_errorbar(aes(ymin = mean-se, ymax=mean+se), 
                width=0.1, 
                position=position_dodge(0.9), 
                color=textcolor) +
  ylab("DV") +
  xlab("Participant Sex") +
  ylim(-5, 5) +
  scale_fill_manual(values = c("#3D66CC", "#892110")) +
  my_theme

```

## Violin Plot

```{r plotcomp-violin}
data %>%
  ggplot(aes(sex, dv, fill = face_sex)) +
  geom_hline(yintercept=0, color=textcolor, size=1) +
  geom_violin(color=textcolor, trim=FALSE, alpha = 0.5) +
  geom_boxplot(color = textcolor, width = 0.25, position = position_dodge(width=0.9)) +
  ylab("DV") +
  xlab("Participant Sex") +
  ylim(-5, 5) + 
  scale_fill_manual(values = c("#3D66CC", "#892110")) +
  my_theme

#ggsave("violin.png", width=10, height = 6)
```

## Split-Violin Plots

To make a split violin plot, first you have to define `geom_split_violin()`. I derived the code from 
[https://stackoverflow.com/questions/35717353/split-violin-plot-with-ggplot2]().

```{r define-split_violin}

GeomSplitViolin <- ggproto(
  "GeomSplitViolin", 
  GeomViolin, 
  draw_group = function(self, data, ..., draw_quantiles = NULL) {
    data <- transform(data, 
                      xminv = x - violinwidth * (x - xmin), 
                      xmaxv = x + violinwidth * (xmax - x))
    grp <- data[1,'group']
    newdata <- plyr::arrange(
      transform(data, x = if(grp%%2==1) xminv else xmaxv), 
      if(grp%%2==1) y else -y
    )
    newdata <- rbind(newdata[1, ], newdata, newdata[nrow(newdata), ], newdata[1, ])
    newdata[c(1,nrow(newdata)-1,nrow(newdata)), 'x'] <- round(newdata[1, 'x']) 
    if (length(draw_quantiles) > 0 & !scales::zero_range(range(data$y))) {
      stopifnot(all(draw_quantiles >= 0), all(draw_quantiles <= 1))
      quantiles <- ggplot2:::create_quantile_segment_frame(data, draw_quantiles)
      aesthetics <- data[rep(1, nrow(quantiles)), setdiff(names(data), c("x", "y")), drop = FALSE]
      aesthetics$alpha <- rep(1, nrow(quantiles))
      both <- cbind(quantiles, aesthetics)
      quantile_grob <- GeomPath$draw_panel(both, ...)
      ggplot2:::ggname("geom_split_violin", 
                       grid::grobTree(GeomPolygon$draw_panel(newdata, ...), quantile_grob))
    } else {
      ggplot2:::ggname("geom_split_violin", GeomPolygon$draw_panel(newdata, ...))
    }
  }
)

geom_split_violin <- function (mapping = NULL, 
                               data = NULL, 
                               stat = "ydensity", 
                               position = "identity", ..., 
                               draw_quantiles = NULL, 
                               trim = TRUE, 
                               scale = "area", 
                               na.rm = FALSE, 
                               show.legend = NA, 
                               inherit.aes = TRUE) {
  layer(data = data, 
        mapping = mapping, 
        stat = stat, 
        geom = GeomSplitViolin, 
        position = position, 
        show.legend = show.legend, 
        inherit.aes = inherit.aes, 
        params = list(trim = trim, 
                      scale = scale, 
                      draw_quantiles = draw_quantiles, 
                      na.rm = na.rm, ...)
        )
}
```

Once you've defined the new geom, you can use `geom_split_violin` pretty much like `geom_violin`. Note how the position of the `geom_boxplot` changes to put the boxplots side-by-side.

```{r plotcomp-split-violin}
data %>%
  ggplot(aes(sex, dv, fill = face_sex)) +
  geom_hline(yintercept=0, color=textcolor, size=1) +
  geom_split_violin(color=textcolor, trim=FALSE, alpha = 0.5) +
  geom_boxplot(color = textcolor, 
               width = 0.25, 
               position = position_dodge(width=0.25)) +
  ylab("DV") +
  xlab("Participant Sex") +
  ylim(-5, 5) + 
  scale_fill_manual(values = c("#3D66CC", "#892110")) +
  my_theme

#ggsave("split_violin.png", width=10, height = 6)
```

