---
title: "Coding Schemes"
author: ~
date: "2017-06-23"
slug: coding-schemes
categories: ["rstats"]
tags: ["r", "coding", "effect", "treatment", "lmer", "mixed effects", "simulation"]
---

```{r setup, message=FALSE}
library(tidyverse)
library(lmerTest)
```

How you choose to code categorical variables changes how you can interpret the intercept and effects of those variables. My favourite [tutorial on coding schemes](http://talklab.psy.gla.ac.uk/tvw/catpred/) explains things in detail. I'm just adding some concrete examples below.

First, I simulated a data frame of 100 raters rating 100 faces each. Female faces get ratings with a mean of 6; male faces get ratings with a mean of 5 (I know, ratings are usually ordinal integers, but let's pretend we used something like a slider). To simulate random effects, both raters and faces have random intercepts with SDs of 1.

```{r, coding-sim-data}

set.seed(555)  # for reproducibility; delete when running simulations

n_raters <- 100
n_faces <- 100

female_mean <- 6
male_mean <- 5

raters <- tibble(
  rater_id = 1:n_raters,
  rater_i = rnorm(n_raters)
)

faces <- tibble(
  face_id = 1:n_faces,
  face_i = rnorm(n_faces),
  face_sex = rep(c("female", "male"), each = n_faces/2)
)

df <- expand.grid(
  face_id = faces$face_id,
  rater_id = raters$rater_id
) %>%
  left_join(faces, by = "face_id") %>%
  left_join(raters, by = "rater_id") %>%
  mutate(
    face_sex_i = ifelse(face_sex=="male", male_mean, female_mean),
    error = rnorm(nrow(.)),
    rating = face_i + rater_i + face_sex_i + error
  )
```

Calculate the means and SDs of the female and male faces.

```{r, echo = FALSE}
sumstat <- df %>%
  group_by(face_sex) %>%
  summarise(
    mean = mean(rating) %>% round(3),
    SD = sd(rating) %>% round(3)
  ) 

sumstat %>% knitr::kable()
```


Always graph your data to confirm you simulated it correctly.

```{r, coding-orig-data}
df %>% 
  ggplot(aes(face_sex, rating)) + 
  geom_violin() +
  geom_boxplot(width=0.2)
```

Recode face sex using treatment, sum, or effect coding.

```{r coding-recode}
df2 <- df %>%
  mutate(
    face_sex.tr = recode(face_sex, "female" = 1, "male" = 0),
    face_sex.sum = recode(face_sex, "female" = -1, "male" = 1),
    face_sex.e = recode(face_sex, "female" = -0.5, "male" = 0.5)
  )
```

Now we analyse the data using each of the 4 styles of coding. I'm just going to show the table of fixed effects.

## Categorical coding
```{r coding-categorical}
m1 <- lmerTest::lmer(rating ~ face_sex + 
                       (1 | face_id) + 
                       (1 + face_sex | rater_id), 
                     data = df2)
```

```{r, echo = FALSE}
sum1 <- summary(m1)
sum1$coefficients %>% round(3) %>% knitr::kable()
```

Note that the intercept coefficient is equal to the female mean (`r sumstat[1,2]`) and the effect of face sex is how much less the male mean is (`r sumstat[2,2]` - `r sumstat[1,2]` = `r sumstat[2,2]-sumstat[1,2]`).

## Treatment coding
```{r coding-treatment}
m.tr <- lmerTest::lmer(rating ~ face_sex.tr + 
               (1 | face_id) + 
               (1 + face_sex.tr | rater_id), 
             data = df2)
```

```{r, echo = FALSE}
sum.tr <- summary(m.tr)
sum.tr$coefficients %>% round(3) %>% knitr::kable()
```

Treatment coding is the same as categorical coding, but gives you more control over what the reference category is. Here, the reference category is `male` and the "treatment" category is `female`, so the intercept coefficient is equal to the male mean (`r sumstat[2,2]`) and the effect of face sex is how much more the female mean is (`r sumstat[1,2]` - `r sumstat[2,2]` = `r sumstat[1,2]-sumstat[2,2]`).

## Sum coding
```{r coding-sum}
m.sum <- lmerTest::lmer(rating ~ face_sex.sum + 
                (1 | face_id) + 
                (1 + face_sex.sum | rater_id), 
              data = df2)
```

```{r, echo = FALSE}
sum.sum <- summary(m.sum)
sum.sum$coefficients %>% round(3) %>% knitr::kable()
```

With sum coding, the intercept coefficient is equal to the overall mean ignoring face sex (i.e.,  (`r sumstat[1,2]` + `r sumstat[2,2]`)/2 = `r (sumstat[1,2] + sumstat[2,2])/2`) and the effect of face sex is how much above and below that each of the two face sexes differ from the mean (i.e., (`r sumstat[1,2]` - `r sumstat[2,2]`)/2 =  `r abs(sumstat[1,2] - sumstat[2,2])/2`).

## Effect coding
```{r coding-effect}
m.e <- lmerTest::lmer(rating ~ face_sex.e + 
              (1 | face_id) + 
              (1 + face_sex.e | rater_id), 
            data = df2)
```

```{r, echo = FALSE}
sum.e <- summary(m.e)
sum.e$coefficients %>% round(3) %>% knitr::kable()
```

With effect coding, the intercept coefficient is the same as sum coding and the effect of face sex is how much the two face sexes differ from each other (i.e., `r sumstat[1,2]` - `r sumstat[2,2]` =  `r abs(sumstat[1,2] - sumstat[2,2])`). Note that this coefficient is double that from the sum coding.
