---
title: "Hidden Solutions"
author: ~
date: "2017-10-21"
slug: hidden-solutions
categories: ["rstats"]
tags: ["r", "pedagogy", "web"]
---

```{r setup, message=FALSE}
library(tidyverse)
```

The [R for Reproducible Scientific Analysis](http://swcarpentry.github.io/r-novice-gapminder/) pages at software carpentry have a really nice interface for hiding and showing solutions to exercises. I've created my own lightweight solution that you can use in any html file, including those made by RMarkdown (e.g., R notebooks).

## Example

Graph the relationship between speed and distance for the `cars` dataset.

<div class="solution"><button>Solution</button>

You can put some text inside the solution, as well as code cunks.

```{r cars}
ggplot(cars, aes(speed, dist)) +
  geom_point(color = "purple") +
  geom_smooth(method = "lm", color = "purple")
```

</div>

Setting this up requires a few lines at the beginning and end of each file, plus surrounding your solutions with a line of html.

## Put this at the beginning of your file

```
<style>
  /* styles for hidden solutions */
  .solution {
    height: 2em;
    overflow-y: hidden;
    padding: 0.5em;
  }
  .solution.open { 
    height: auto; 
    background-color: rgba(0, 0, 0, 0.1);
    border-radius: 5px;
  }
  .solution button {
    height: 1.5em;
    margin-bottom: 0.5em;
  }
  .solution pre.sourceCode {
    border-color: green;
  }
</style>
```

If you're using [RMarkdown Websites](http://rmarkdown.rstudio.com/rmarkdown_websites.html), you can just put these lines of css into an external stylesheet linked in your `_site.yml` file (e.g., [rguppies.css](rguppies.css)).

## Put this at the end of your file

```
<script>
  window.onload = function(){
    var buttons = document.getElementsByTagName("button");
    for (var i = 0; i < buttons.length; i++) {
      buttons[i].onclick = function() {
        var cl = this.parentElement.classList;
        if (cl.contains('open')) {
          cl.remove("open");
        } else {
          cl.add("open");
        }
      }
    }
  }
</script>
```

If you're using [RMarkdown Websites](http://rmarkdown.rstudio.com/rmarkdown_websites.html), you can just put this script into an external footer or script file linked in your `_site.yml` file (e.g., [rguppies.js](rguppies.js)).

## Surround your hidden solutions like this

```
<div class="solution"><button>Solution</button>

PUT YOUR SOLUTION HERE (including r chunks)

</div>
```
You can change the text on the button to something else (e.g., `<button>View the Answer</button>`).

Let me know if you have any suggestions or find this useful.