---
title: Simulating Multiple Vectors
author: ~
date: '2018-12-22'
slug: rnorm_multi
categories: ["rstats"]
tags: ["R", "correlated data", "simulation", "faux"]
---

```{r load-libraries, message = FALSE}
library(tidyverse)
library(faux)
```

I'm working on a package for simulations called [faux](https://github.com/debruine/faux). (Update: faux is now on CRAN!)

The first function, `rnorm_multi`, makes multiple normally distributed vectors with specified relationships and takes the following arguments:

* `n` = the number of samples required (required)
* `vars` = the number of variables to return (default = `3`)
* `cors` = the correlations among the variables (can be a single number, vars\*vars matrix, vars\*vars vector, or a vars\*(vars-1)/2 vector; default = `0`)
* `mu` = a vector giving the means of the variables (numeric vector of length 1 or vars; default = `0`)
* `sd` = the standard deviations of the variables (numeric vector of length 1 or vars; default = `1`)
* `varnames` = optional names for the variables (string vector of length vars; default = `NULL`)
* `empirical` = logical. If true, mu, sd and cors specify the empirical not population mean, sd and covariance (default = `FALSE`)


## Example 1

The following example creates a 100-row dataframe of 3 columns names `A`, `B`, and `C`, with means = 0, SDs = 1, and where r~AB~ = 0.2, r~AC~ = -0.5, and r~BC~ = 0.5.

```{r}
ex1 <- rnorm_multi(100, 3, c(0.2, -0.5, 0.5), varnames=c("A", "B", "C"))
```

### Correlation Matrix of Sample Data

`r cor(ex1) %>% knitr::kable()`

## Example 2

The following example calculates the correlation matrix, means, and SDs from the `iris` dataset and uses them to simulate a dataset of 100 rows with the same parameters.

```{r}

dat <- select_if(iris, is.numeric)

iris_sim <- rnorm_multi(
  n = 100, 
  vars = ncol(dat), 
  r = cor(dat),
  mu = summarise_all(dat, mean) %>% t(), 
  sd = summarise_all(dat, sd) %>% t(), 
  varnames = names(dat)
)

```

### Correlation Matrix of Original Data

`r cor(dat) %>% knitr::kable()`

### Correlation Matrix of Sample Data

`r cor(iris_sim) %>% knitr::kable()`