---
title: "Testing interactive functions"
author: ~
date: "2020-07-31"
slug: interactive-test
categories: ["rstats"]
tags: ["R", "interactive", "test", "testthat"]
---

I'm a huge fan of unit tests, but it's tricky to test interactive functions where the user needs to enter input before the function can progress. I used to test them manually, which is incredibly tedious and time-consuming. So I ended up not testing interative functions very thoroughly. I found a [post on Stack Overflow](https://stackoverflow.com/questions/41372146/test-interaction-with-users-in-r-package) with a useful answer by [znk](https://stackoverflow.com/users/2752888/znk). I've expanded their answer into a full example of a unit test for an interactive function.

```{r setup}
library(testthat)
```

# Set up the function

Your function needs to use `readLines` to get interactive input and take an argument for the connection (`con`). The default value for the connection should be the same as its default value for `readLines`, which is `stdin()` (the terminal). You can't use `readline`, which only supports connection to the terminal.

This function displays a prompt and a list of valid options. If your response isn't in the list, it will repeat the prompt until it is.

```{r}
ask_opts <- function(prompt, opts = NULL, con = stdin()) {
  # display prompt and options
  optlist <- paste(opts, collapse = "|")
  prompt_opt <- paste0(prompt, " [", optlist, "]\n")
  cat(prompt_opt)
  input <- readLines(con = con, n = 1)
  
  # repeat if input is not in opts
  if (!is.null(opts) & !input %in% opts) {
    input <- ask_opts(prompt, opts, con)
  }

  input
}
```


## Set up the Test

You need to create a file containing the input you want to send to the function, one input per line. I want to answer the first time with something not in the option list, then the second time with something that is in the option list.

```{r}
# set up interactive answers
f <- file()
lines <- c("echidna", "ferret")
ans <- paste(lines, collapse = "\n")
write(ans, f)
```

Then run your interactive function, setting the connection to your file. Run it inside `capture_output_lines` if you want to test the prompts and not just the output. Close the file when you are done with it.

```{r}
prompt <- "What is your favourite animal?"
opts <- c("cat", "dog", "ferret")

output_prompts <- capture_output_lines({
  result <- ask_opts(prompt, opts, f)
})

close(f) # close the file
```

Now you can run your tests

```{r}
txt <- "What is your favourite animal? [cat|dog|ferret]"
expect_equal(result, "ferret")
expect_equal(output_prompts, rep(txt, 2))
```

## Without a new argument

What if you don't want to change the arguments to your function to add a connection? You can set the connection in the options and test for it in the function, defaulting to `stdin()`. For example:

```{r}
ask_opts <- function(prompt, opts = NULL) {
  # set up connection, default to stdin() if not set
  con <- getOption("ask_opts.con", stdin())
  
  # display prompt and options
  optlist <- paste(opts, collapse = "|")
  prompt_opt <- paste0(prompt, " [", optlist, "]\n")
  cat(prompt_opt)
  input <- readLines(con = con, n = 1)
  
  # repeat if input is not in opts
  if (!is.null(opts) & !input %in% opts) {
    input <- ask_opts(prompt, opts)
  }

  input
}
```

Then you just need to set this option before you run the interactive function in your testing environment. Make sure to reset it to `stdin()` when you're done.

```{r}
test_that("interactive", {
  # set up interactive answers
  f <- file()
  lines <- c("maybe", "y")
  ans <- paste(lines, collapse = "\n")
  write(ans, f)
  
  options("ask_opts.con" = f) # set connection option 
  
  # run interactive function
  prompt <- "Was this helpful?"
  opts <- c("y", "n")
  
  output_prompts <- capture_output_lines({
    result <- ask_opts(prompt, opts)
  })
  
  close(f) # close the file
  options("ask_opts.con" = stdin()) # reset connection option
  
  # tests
  txt <- "Was this helpful? [y|n]"
  expect_equal(result, "y")
  expect_equal(output_prompts, rep(txt, 2))
})
```