---
title: "PCA"
author: ~
date: "2017-06-26"
slug: pca
categories: ["rstats"]
tags: ["r", "pca", "princomp", "psych", "factanal"]
---

[Download the Rmd notebook for this example](pca.Rmd)

Putting together this page made me realise I still don't know anything about PCA and factor analysis.

I use the <code>psych</code> package for SPSS-style PCA.

```{r load-libraries, message = FALSE}
library(tidyverse)
library(psych)
library(viridis)
```

First, I'll simulate some data with an underlying structure of three factors.

```{r pca-sim-data}

set.seed(444)  # for reproducibility; delete when running simulations

a <- rnorm(100, 0, 1)
b <- rnorm(100, 0, 1)
c <- rnorm(100, 0, 1)

df <- data.frame(
  id = seq(1,100),
  a1 = a + rnorm(100, 0, 1),
  a2 = a + rnorm(100, 0, .8),
  a3 = a + rnorm(100, 0, .6),
  a4 = -a + rnorm(100, 0, .4),
  b1 = b + rnorm(100, 0, 1),
  b2 = b + rnorm(100, 0, .8),
  b3 = b + rnorm(100, 0, .6),
  b4 = -b + rnorm(100, 0, .4),
  c1 = c + rnorm(100, 0, 1),
  c2 = c + rnorm(100, 0, .8),
  c3 = c + rnorm(100, 0, .6),
  c4 = -c + rnorm(100, 0, .4)
)
```

Select just the columns you want for your PCA. You can visualise their correlations with <code>cor()</code> and <code>ggplot()</code>.

```{r pca-cor-vars}
traits <- df %>% select(-id) 

traits %>%
  cor() %>%
  as.data.frame() %>%
  mutate(var1 = rownames(.)) %>%
  gather("var2", "value", a1:c4) %>%
  mutate(var1 = factor(var1), var1 = factor(var1, levels = rev(levels(var1)))) %>%
  ggplot(aes(var2, var1, fill=value)) +
  geom_tile() +
  scale_x_discrete(position = "top") +
  xlab("") + ylab("") +
  scale_fill_viridis(limits=c(-1, 1))

```

Determine the number of factors to extract. Here I use the SPSS-style default criterion of  Eigenvalues > 1
```{r pca-nfactors}
ev <- eigen(cor(traits));
nfactors <- length(ev$values[ev$values > 1]);
nfactors
```


## Principal components analysis (SPSS-style)

### principal(rotation = "none")
```{r}
traits.principal <- principal(traits, nfactors=nfactors, rotate="none", scores=TRUE)
traits.principal
scores.principal <- traits.principal$scores
```

```{r pca-principal-none, fig.width=7, fig.height=3}
cor(scores.principal, traits) %>%
  as.data.frame() %>%
  mutate(var1 = rownames(.)) %>%
  gather("var2", "value", a1:c4) %>%
  mutate(var1 = factor(var1), var1 = factor(var1, levels = rev(levels(var1)))) %>%
  ggplot(aes(var2, var1, fill=value)) +
  geom_tile() +
  scale_x_discrete(position = "top") +
  xlab("") + ylab("") +
  scale_fill_viridis(limits=c(-1, 1))
```

### principal(rotation = "varimax")
```{r}
traits.varimax <- principal(traits, nfactors=nfactors, rotate="varimax", scores=TRUE)
traits.varimax
scores.varimax <- traits.varimax$scores
```

```{r pca-principal-varimax, fig.width=7, fig.height=3}
cor(scores.varimax, traits) %>%
  as.data.frame() %>%
  mutate(var1 = rownames(.)) %>%
  gather("var2", "value", a1:c4) %>%
  mutate(var1 = factor(var1), var1 = factor(var1, levels = rev(levels(var1)))) %>%
  ggplot(aes(var2, var1, fill=value)) +
  geom_tile() +
  scale_x_discrete(position = "top") +
  xlab("") + ylab("") +
  scale_fill_viridis(limits=c(-1, 1))
```

## Here are some other functions for PCA/Factor Analysis

### princomp()
```{r}
traits.princomp <- princomp(traits)
traits.princomp$loadings[,1:nfactors]
scores.princomp <- traits.princomp$scores[,1:nfactors]
```

```{r pca-princomp, fig.width=7, fig.height=3}
cor(scores.princomp, traits) %>%
  as.data.frame() %>%
  mutate(var1 = rownames(.)) %>%
  gather("var2", "value", a1:c4) %>%
  mutate(var1 = factor(var1), var1 = factor(var1, levels = rev(levels(var1)))) %>%
  ggplot(aes(var2, var1, fill=value)) +
  geom_tile() +
  scale_x_discrete(position = "top") +
  xlab("") + ylab("") +
  scale_fill_viridis(limits=c(-1, 1))
```

### factanal(rotation = "none")
```{r}
traits.fa <- factanal(traits, nfactors, rotation="none", scores="regression")
print(traits.fa, digits=2, cutoff=0, sort=FALSE)
scores.fa <- traits.fa$scores
```

```{r pca-factanal-none, fig.width=7, fig.height=3}
cor(scores.fa, traits) %>%
  as.data.frame() %>%
  mutate(var1 = rownames(.)) %>%
  gather("var2", "value", a1:c4) %>%
  mutate(var1 = factor(var1), var1 = factor(var1, levels = rev(levels(var1)))) %>%
  ggplot(aes(var2, var1, fill=value)) +
  geom_tile() +
  scale_x_discrete(position = "top") +
  xlab("") + ylab("") +
  scale_fill_viridis(limits=c(-1, 1))
```

### factanal(rotation = "varimax")
```{r}
traits.fa.vm <- factanal(traits, nfactors, rotation="varimax", scores="regression")
print(traits.fa.vm, digits=2, cutoff=0, sort=FALSE)
scores.fa.vm <- traits.fa.vm$scores
```

```{r pca-factanal-varimax, fig.width=7, fig.height=3}
cor(scores.fa.vm, traits) %>%
  as.data.frame() %>%
  mutate(var1 = rownames(.)) %>%
  gather("var2", "value", a1:c4) %>%
  mutate(var1 = factor(var1), var1 = factor(var1, levels = rev(levels(var1)))) %>%
  ggplot(aes(var2, var1, fill=value)) +
  geom_tile() +
  scale_x_discrete(position = "top") +
  xlab("") + ylab("") +
  scale_fill_viridis(limits=c(-1, 1))
```

## How do they compare?

Here, I'll plot the absolute value of all the correlations (since the sign on factors/PCs is arbitrary).

The functions principal(rotation = "varimax") and factanal(rotation = "varimax") are nearly (but not perfectly) identical.
```{r pca-compare-all, fig.width=7, fig.height=6}
scores.fa <- traits.fa$scores
colnames(scores.principal) <- c("principal() 1", "principal() 2", "principal() 3")
colnames(scores.varimax) <- c("principal(vm) 1", "principal(vm) 2", "principal(vm) 3")
colnames(scores.princomp) <- c("princomp() 1", "princomp() 2", "princomp() 3")
colnames(scores.fa) <- c("factanal() 1", "factanal() 2", "factanal() 3")
colnames(scores.fa.vm) <- c("factanal(vm) 1", "factanal(vm) 2", "factanal(vm) 3")

cbind(
  scores.princomp,
  scores.principal,
  scores.fa,
  scores.varimax,
  scores.fa.vm
) %>% 
  cor() %>%
  as.data.frame() %>%
  mutate(var1 = rownames(.)) %>%
  gather("var2", "value", 1:15) %>%
  mutate(var1 = factor(var1), var1 = factor(var1, levels = rev(levels(var1)))) %>%
  mutate(value = abs(value)) %>%
  ggplot(aes(var2, var1, fill=value)) +
  geom_tile() +
  scale_x_discrete(position = "top") +
  xlab("") + ylab("") +
  scale_fill_viridis(limits=c(0, 1)) +
  geom_hline(yintercept = 3.5, color="white") +
  geom_hline(yintercept = 6.5, color="white") +
  geom_hline(yintercept = 9.5, color="white") +
  geom_hline(yintercept = 12.5, color="white") +
  geom_vline(xintercept = 3.5, color="white") +
  geom_vline(xintercept = 6.5, color="white") +
  geom_vline(xintercept = 9.5, color="white") +
  geom_vline(xintercept = 12.5, color="white") +
  theme(axis.text.x=element_text(angle=90,hjust=1))
```
