---
title: Simulating Mixed Effects
author: ~
date: "2018-12-26"
slug: simulating-mixed-effects
categories: ["rstats"]
tags: ["R", "simulation", "lmer", "lme4"]
---

[Updated 2019-01-09]

I'm going to walk through one example of simulating a dataset with random effects. I'll generate data for a Stroop task where people (`subjects`) say the colour of  colour words (`stimuli`) shown in each of two versions (`congruent` and `incongruent`). Subjects are in one of two conditions (`hard` and `easy`). The dependent variable (`DV`) is reaction time. 

* congruent: <span style="color: red;">RED</span>, <span style="color: orange;">ORANGE</span>, <span style="color: #ead21e;">YELLOW</span>, <span style="color: green;">GREEN</span>, <span style="color: blue;">BLUE</span>, <span style="color: purple;">PURPLE</span>
* incongruent: <span style="color: #ead21e;">RED</span>, <span style="color: green;">ORANGE</span>, <span style="color: blue;">YELLOW</span>, <span style="color: purple;">GREEN</span>, <span style="color: red;">BLUE</span>, <span style="color: orange;">PURPLE</span>

We expect people to have faster reaction times for congruent stimuli than incongruent stimuli (main effect of version) and to be faster in the easy condition than the hard condition (main effect of condition). We'll look at some different interaction patterns below. 

## Setup

We'll use the `tidyverse` to manipulate data frames and `lmerTest` (which includes `lmer`) to run the mixed effects models. I also like to set the `scipen` and `digits` options to get rid of scientific notation in lmer output.

When you're simulating data, you should start your script by setting a seed. You can use any number you like, this just makes sure that you get the same results every time you run the script. (Thanks for reminding me to add and explain this, [Tim Morris](https://twitter.com/tmorris_mrc)!)

```{r setup, message=FALSE}
library(tidyverse)
library(lmerTest)
options("scipen"=10, "digits"=4)
set.seed(8675309) # Jenny, I've got your number
```

<p class="warning">If you are running repeated simulations (e.g., for a power calculation), make sure you never use `set.seed` inside of a function that creates random numbers, or the function will always give you the exact same numbers.</p>

## Simulate Random Effects

### Subjects

First, we need to generate a sample of subjects. Each subject will have slightly faster or slower reaction times on average; this is their random intercept (`sub_i`). We'll model it from a normal distribution with a mean of 0 and SD of 100ms. 

We also add between-subject variables here. Each subject is in only one condition, so assign half `easy` and half `hard`. Set the number of subjects as `sub_n` at the beginning so you can change this in the future with only one edit.


```{r sim-subject}
sub_n  <- 200 # number of subjects in this simulation
sub_sd <- 100 # SD for the subjects' random intercept

sub <- tibble(
  sub_id = 1:sub_n,
  sub_i  = rnorm(sub_n, 0, sub_sd), # random intercept
  sub_cond = rep(c("easy","hard"), each = sub_n/2) # between-subjects factor
)
```

<p class="info">If you already have [pilot data](/data/pilot_data.csv), you can estimate a realistic SD for the subjects' random intercepts using the following code.</p>

```{r calc-sub-sd, eval = FALSE}

pilot_data <- read_csv("pilot_data.csv")
pilot_mod <- lmer(dv ~ 1 + (1 | sub_id) + (1 | stim_id), data = pilot_data)

sub_sd <- VarCorr(pilot_mod) %>% 
  as.data.frame() %>%
  filter(grp == "sub_id") %>% 
  pull(sdcor)

```

I like to check my simulations at every step with a graph. We expect subjects in  `hard` and `easy` conditions to have approximately equal intercepts.

```{r plot-subject, fig.cap="Double-check subject intercepts"}
ggplot(sub, aes(sub_i, color = sub_cond)) +
  geom_density()
```

### Stimuli

Now, we generate a sample of stimuli. Each stimulus will have slightly faster or slower reaction times on average; this is their random intercept (`stim_i`). We'll model it from a normal distribution with a mean of 0 and SD of 50ms (it seems reasonable to expect less variability between words than people for this task). Stimulus version (`congruent` vs `incongruent`) is a within-stimulus variable, so we don't need to add it here.

```{r sim-stimuli}

stim_n  <- 50 # number of stimuli in this simulation
stim_sd <- 50 # SD for the stimuli's random intercept

stim <- tibble(
  stim_id = 1:stim_n,
  stim_i = rnorm(stim_n, 0, stim_sd) # random intercept
)

```

<p class="info">If you want to simulate data using the exact stimuli you already have pilot data for, you can calculate their random intercepts using the following code and use that instead of generating a random sample with the same SD.</p>

```{r exact-stim, eval = FALSE}

pilot_data <- read_csv("pilot_data.csv")
pilot_mod <- lmer(dv ~ 1 + (1 | sub_id) + (1 | stim_id), data = pilot_data)

stim <- ranef(pilot_mod)$stim_id %>%
  as_tibble(rownames = 'stim_id') %>%
  rename(stim_i = `(Intercept)`)

```


Plot the random intercepts to double-check they look like you expect.

```{r plot-stimuli, fig.cap="Double-check stimulus intercepts"}
ggplot(stim, aes(stim_i)) +
  geom_density()
```

### Trials

Now we put the subjects and stimuli together. In this study, all subjects respond to all stimuli in both upright and inverted versions, but subjects are in only one condition. The function `expand.grid` gives you a data frame with all possible combinations of the arguments. Add the data specific to each subject and stimulus by left joining the `sub` and `stim` data frames.

```{r expand-grid}
trials <- expand.grid(
  sub_id = sub$sub_id, # get subject IDs from the sub data table
  stim_id = stim$stim_id, # get stimulus IDs from the stim data table
  stim_version = c("congruent", "incongruent") # all subjects see both congruent and incongruent versions of all stimuli
) %>%
  left_join(sub, by = "sub_id") %>% # includes the intercept and conditin for each subject
  left_join(stim, by = "stim_id")   # includes the intercept for each stimulus
```

`r trials %>% filter(sub_id < 3, stim_id < 3) %>% knitr::kable(caption="Subject- and stimulus-specific random intercepts for 2 subjects and 2 stimuli")`

## Calculate DV

Now we can calculate the DV by adding together an overall intercept (mean RT for all trials), the subject-specific intercept, the stimulus-specific intercept, the effect of subject condition, the interaction between condition and version (set to 0 for this first example), the effect of stimulus version, and an error term. 

### Simulating Fixed Effects

We set these effects in raw units (ms) and [effect-code](/posts/coding-schemes.html) the subject condition and stimulus version. It's usually easiest to interpret if you recode the level that you predict will be larger as +0.5 and the level you predict will be smaller as -0.5. So when we set the effect of subject condition (`sub_cond_eff`) to 50, that means the average difference between the easy and hard condition is 50ms. `Easy` is effect-coded as -0.5 and `hard` is effect-coded as +0.5, which means that trials in the easy condition have -0.5 \* 50ms (i.e., -25ms) added to their reaction time, while trials in the hard condition have +0.5 \* 50ms (i.e., +25ms) added to their reaction time.

```{r sim-dv-vars}
# set variables to use in calculations below
grand_i          <- 400 # overall mean DV
sub_cond_eff     <- 50  # mean difference between conditions: hard - easy
stim_version_eff <- 50  # mean difference between versions: incongruent - congruent
cond_version_ixn <-  0  # interaction between version and condition
error_sd         <- 25  # residual (error) SD
```

<p class="info">We set the error SD fairly low here so that it's easier to see how the parameters we set map onto the analysis output. You can calculate a realistic error SD from pilot data with the code below.</p>

```{r calc-error-sd, eval = FALSE}
pilot_data <- read_csv("pilot_data.csv")
pilot_mod <- lmer(dv ~ 1 + (1 | sub_id) + (1 | stim_id), data = pilot_data)

error_sd <- VarCorr(pilot_mod) %>% 
  as.data.frame() %>%
  filter(grp == "Residual") %>% 
  pull(sdcor)
```

The code chunk below effect-codes the condition and version factors (important for the analysis below), generates an error term for each trial, and generates the DV.

```{r sim-dv}
dat <- trials %>%
  mutate(
    # effect-code subject condition and stimulus version
    sub_cond.e = recode(sub_cond, "hard" = -0.5, "easy" = +0.5),
    stim_version.e = recode(stim_version, "congruent" = -0.5, "incongruent" = +0.5),
    # calculate error term (normally distributed residual with SD set above)
    err = rnorm(nrow(.), 0, error_sd),
    # calculate DV from intercepts, effects, and error
    dv = grand_i + sub_i + stim_i + err +
         (sub_cond.e * sub_cond_eff) + 
         (stim_version.e * stim_version_eff) + 
         (sub_cond.e * stim_version.e * cond_version_ixn) # in this example, this is always 0 and could be omitted
  )

```

As always, graph to make sure you've simulated the general pattern you expected.

```{r plot-dv, fig.cap="Double-check the simulated pattern"}
ggplot(dat, aes(sub_cond, dv, color = stim_version)) +
  geom_hline(yintercept = grand_i) +
  geom_violin(alpha = 0.5) +
  geom_boxplot(width = 0.2, position = position_dodge(width = 0.9))
```

Let's take a concrete example.

`r dat %>% filter(sub_id == 1, stim_id == 1, stim_version == "congruent") %>% knitr::kable(caption="Subject 1's reaction time to stimulus 1 in the congruent condition")`

The DV is equal to the grand intercept (`r grand_i`) plus the subject's intercept (`r sub$sub_i[1]`) plus the stimulus' intercept (`r stim$stim_i[1]`) plus the effect code for the subject's condition (`r sub$sub_cond[1]` = -0.5) times the effect size for subject condition (`r sub_cond_eff`) plus the effect code for stimulus version (congruent = -0.5) times the effect size for stimulus version (`r stim_version_eff`) plus some random error (`r dat$err[1]`).

`r grand_i` + `r sub$sub_i[1]` + `r stim$stim_i[1]` + (-0.5 \* `r sub_cond_eff`) + (-0.5 \* `r stim_version_eff`) + `r dat$err[1]` = `r dat$dv[1]`




```{r calc-dv, include = F}
# function for calculating the difference between the means of two groups
mean_diff <- function(dat, dv, groupby) {
  dat %>%
    select(grp = one_of(groupby), dv = one_of(dv)) %>%
    group_by(grp) %>%
    summarise(m = mean(dv)) %>%
    ungroup() %>%
    spread(grp, m) %>%
    rename_at(1:2, ~ c("hard", "easy")) %>%
    mutate(eff = hard - easy) %>%
    pull(eff) %>%
    round(1)
}

sim_grand_i <- mean(dat$dv) %>% round(1)
sim_cond_eff <- mean_diff(dat, "dv", "sub_cond")
sim_version_eff <- mean_diff(dat, "dv", "stim_version")
```

In this simulated dataset, the grand intercept is `r sim_grand_i`, the mean difference between subject conditions is `r sim_cond_eff`, and the mean difference between stimulus versions is `r sim_version_eff`.

### Simulating Interactions

If you want to simulate an interaction, it can be tricky to figure out what to set the main effects and interaction effect to. It can be easier to think about the simple main effects for each cell. Create four new variables and set them to the deviations from the overall mean you'd expect for each condition (so they should add up to 0). Here, we're simulating a small effect of version in the hard condition (50ms difference) and double that effect of version in the easy condition (100ms difference).

```{r sim-simple-main-effects}
# set variables to use in calculations below
grand_i    <- 400
hard_congr <- -25
hard_incon <- +25
easy_congr <- -50
easy_incon <- +50
error_sd   <-  25
```

Use the code below to transform the simple main effects above into main effects and interactions for use in the equations below.

````{r sim-effect-calc}
# calculate main effects and interactions from simple effects above

# mean difference between easy and hard conditions
sub_cond_eff     <- (easy_congr + easy_incon)/2 -
                    (hard_congr + hard_incon)/2
# mean difference between incongruent and congruent versions
stim_version_eff <- (hard_incon + easy_incon)/2 - 
                    (hard_congr + easy_congr)/2  
# interaction between version and condition
cond_version_ixn <- (easy_incon - easy_congr) -
                    (hard_incon - hard_congr) 
                    
```

Then generate the DV the same way we did above, but also add the interaction effect multiplied by the effect-coded subject condition and stimulus version.

```{r sim-ixn}
dat <- trials %>%
  mutate(
    # effect-code subject condition and stimulus version
    sub_cond.e = recode(sub_cond, "hard" = -0.5, "easy" = +0.5),
    stim_version.e = recode(stim_version, "congruent" = -0.5, "incongruent" = +0.5),
    # calculate error term (normally distributed residual with SD set above)
    err = rnorm(nrow(.), 0, error_sd),
    # calculate DV from intercepts, effects, and error
    dv = grand_i + sub_i + stim_i + err +
         (sub_cond.e * sub_cond_eff) + 
         (stim_version.e * stim_version_eff) + 
         (sub_cond.e * stim_version.e * cond_version_ixn)
  )

```

```{r plot-ixn, fig.cap="Double-check the interaction between condition and version"}
ggplot(dat, aes(sub_cond, dv, color = stim_version)) +
  geom_hline(yintercept = grand_i) +
  geom_violin(alpha = 0.5) +
  geom_boxplot(width = 0.2, position = position_dodge(width = 0.9))
```

Let's look at subject 1's reaction time to stimulus 1 in the congruent condition in more detail again.

`r dat %>% filter(sub_id == 1, stim_id == 1, stim_version == "congruent") %>% knitr::kable(caption="Subject 1's reaction time to stimulus 1 in the congruent condition")`

The DV is equal to the grand intercept (`r grand_i`) plus the subject's intercept (`r sub$sub_i[1]`) plus the stimulus' intercept (`r stim$stim_i[1]`) plus the effect code for the subject's condition (`r sub$sub_cond[1]` = -0.5) times the effect size for subject condition (`r sub_cond_eff`) plus the effect code for stimulus version (congruent = -0.5) times the effect size for stimulus version(`r stim_version_eff`) plus the effect code for the subject's condition (`r sub$sub_cond[1]` = -0.5) times the effect code for stimulus version (congruent = -0.5) times the effect size for the interaction (`r cond_version_ixn`) plus some random error (`r dat$err[1]`).

`r grand_i` + `r sub$sub_i[1]` + `r stim$stim_i[1]` + (-0.5 \* `r sub_cond_eff`) + (-0.5 \* `r stim_version_eff`) + (-0.5 \* -0.5 \* `r cond_version_ixn`) + `r dat$err[1]` = `r dat$dv[1]`



```{r table-ixn, results='asis'}
group_by(dat, sub_cond, stim_version) %>%
  summarise(m = mean(dv) - grand_i %>% round(1)) %>%
  ungroup() %>%
  spread(stim_version, m) %>%
  knitr::kable()
```

## Analysis

New we will run a linear mixed effects model with `lmer` and look at the summary. 

```{r lmer}
mod <- lmer(dv ~ sub_cond.e * stim_version.e +
              (1 | sub_id) + 
              (1 | stim_id),
            data = dat)

mod.sum <- summary(mod)

mod.sum
```

### Sense Checks

First, check that your groups make sense (`mod.sum$ngrps`). 

* The number of obs should be the total number of trials analysed. 
* `sub_id` should be what we set `sub_n` to above (`r sub_n`).
* `stim_id` should be what we set `stim_n` to above (`r stim_n`).

```{r mod-ngrps, echo = FALSE}
mod.sum$ngrps
```

Next, look at the random effects (`mod.sum$varcor`). 

* The SD for `sub_id` should be near the `sub_sd` of `r sub_sd`.
* The SD for `stim_id` should be near the `stim_sd` off `r stim_sd`. 
* The residual SD should be near the `error_sd` of `r error_sd`.

```{r mod-varcor, echo = FALSE}
mod.sum$varcor
```

Finally, look at the fixed effects (`mod.sum$coefficients`). 

* The estimate for the Intercept should be near the `grand_i` of `r grand_i`. 
* The main effect of `sub_cond.e` should be near what we calculated for `sub_cond_eff` (`r sub_cond_eff`). 
* The main effect of `stim_version.e` should be near what we calculated for `stim_version_eff` (`r stim_version_eff`)
* The interaction between `sub_cond.e`:`stim_version.e` should be near what we calculated for `cond_version_ixn` (`r cond_version_ixn`).

```{r mod-coef, echo = FALSE}
mod.sum$coefficients
```

### Random effects

Plot the subject intercepts from our code above (`sub$sub_i`) against the subject intercepts calculcated by `lmer` (`ranef(mod)$sub_id`).

```{r plot-sub-ranef, fig.cap = "Compare simulated subject random intercepts to those from the model"}
ranef(mod)$sub_id %>%
  as_tibble(rownames = "sub_id") %>%
  rename(mod_sub_i = `(Intercept)`) %>%
  mutate(sub_id = as.integer(sub_id)) %>%
  left_join(sub, by = "sub_id") %>%
  ggplot(aes(sub_i,mod_sub_i)) +
  geom_point() +
  geom_smooth(method = "lm") +
  xlab("Simulated random intercepts (sub_i)") +
  ylab("Modeled random intercepts")
```

Plot the stimulus intercepts from our code above (`stim$stim_i`) against the stimulus intercepts calculcated by `lmer` (`ranef(mod)$stim_id`).

```{r plot-stim-ranef, fig.cap = "Compare simulated stimulus random intercepts to those from the model"}
ranef(mod)$stim_id %>%
  as_tibble(rownames = "stim_id") %>%
  rename(mod_stim_i = `(Intercept)`) %>%
  mutate(stim_id = as.integer(stim_id)) %>%
  left_join(stim, by = "stim_id") %>%
  ggplot(aes(stim_i,mod_stim_i)) +
  geom_point() +
  geom_smooth(method = "lm") +
  xlab("Simulated random intercepts (stim_i)") +
  ylab("Modeled random intercepts")
```


## Function

You can put the code above in a function so you can run it more easily and change the parameters. I removed the plot and set the argument defaults to the same as the example above, but you can set them to other patterns.

```{r sim-function}
sim_lmer <- function( sub_n = 200,
                      sub_sd = 100,
                      stim_n = 50,
                      stim_sd = 50,
                      grand_i = 400,
                      hard_congr = -25,
                      hard_incon = +25,
                      easy_congr = -50,
                      easy_incon = +50,
                      error_sd = 25) {
  sub <- tibble(
    sub_id = 1:sub_n,
    sub_i  = rnorm(sub_n, 0, sub_sd),
    sub_cond = rep(c("hard","easy"), each = sub_n/2)
  )
  
  stim <- tibble(
    stim_id = 1:stim_n,
    stim_i = rnorm(stim_n, 0, stim_sd)
  )

  # mean difference between easy and hard conditions
  sub_cond_eff     <- (easy_congr + easy_incon)/2 -
                      (hard_congr + hard_incon)/2
  # mean difference between incongruent and congruent versions
  stim_version_eff <- (hard_incon + easy_incon)/2 - 
                      (hard_congr + easy_congr)/2  
  # interaction between version and condition
  cond_version_ixn <- (easy_incon - easy_congr) -
                      (hard_incon - hard_congr) 
  
  dat <- expand.grid(
    sub_id = sub$sub_id,
    stim_id = stim$stim_id,
    stim_version = c("congruent", "incongruent")
  ) %>%
    left_join(sub, by = "sub_id") %>%
    left_join(stim, by = "stim_id") %>%
    mutate(
      # effect-code subject condition and stimulus version
      sub_cond.e = recode(sub_cond, "hard" = -0.5, "easy" = +0.5),
      stim_version.e = recode(stim_version, "congruent" = -0.5, "incongruent" = +0.5),
      # calculate error term (normally distributed residual with SD set above)
      err = rnorm(nrow(.), 0, error_sd),
      # calculate DV from intercepts, effects, and error
      dv = grand_i + sub_i + stim_i + err +
           (sub_cond.e * sub_cond_eff) + 
           (stim_version.e * stim_version_eff) + 
           (sub_cond.e * stim_version.e * cond_version_ixn)
    )
  
  mod <- lmer(dv ~ sub_cond.e * stim_version.e +
                (1 | sub_id) + 
                (1 | stim_id),
              data = dat)
  
  mod.sum <- summary(mod)
  
  return(mod.sum)
}
```

Run the function with the default values.

```{r sim-lmer-default}
sim_lmer()
```

Try changing some variables to simulate null effects.

```{r sim-lmer-changes}
sim_lmer(hard_congr = 0,
         hard_incon = 0,
         easy_congr = 0,
         easy_incon = 0)
```

