---
title: "scale()"
author: "Lisa DeBruine"
date: "2017-06-22"
slug: scale
categories: ["rstats"]
tags: ["R", "scale"]
---

<p>You can use <code>scale()</code> to center and/or scale (i.e., Z-score) a vector of numbers.</p>

Z-score a list of numbers
```{r}
x <- c(10, 12, 14, 16, 18)
scale(x)
```

However, the result contains the mean and SD. 
  This can cause problems if you want to assign it to a new column in a data frame, 
  which you can fix using <code>as.vector()</code>
```{r} 
as.vector(scale(x))
```

I find it more straightforward to just use the equation for a Z-score
```{r}
( x - mean(x) ) / sd(x)
```

You can just center the numbers without scaling.
```{r}
as.vector(scale(x, center=TRUE, scale=FALSE))
( x - mean(x) )
```

Scaling without centering divides numbers by their root mean square.
```{r}
as.vector(scale(x, center=FALSE, scale=TRUE))
x / sqrt(sum(x^2)/(length(x)-1))
```
  
Set the scale to a number to divide by that number
```{r}
as.vector(scale(x, center=FALSE, scale=3))
x / 3
```
  
Create new columns in a dataframe with the scaled or centered variable
```{r}
suppressMessages( library(tidyverse) )
df <- data.frame(id = seq(1,5), x = x)
df.s <- df %>%
  mutate(
    x.s = as.vector(scale(x)),
    x.c = as.vector(scale(x, scale=F)),
    x.z = (x - mean(x)) / sd(x)
  )
df.s
```

